---
title: "Fetal Health Classification Using Machine Learning"
author: "Yixian Cai | Jiali Wu"
date: "2026-05-08"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages, include=FALSE}
# ── Data wrangling & reporting ──────────────────────────────
library(tidyverse)
library(knitr)
library(kableExtra)

# ── Machine learning ────────────────────────────────────────
library(caret)
library(randomForest)
library(gbm)
library(nnet)
library(e1071)

# ── Visualisation & evaluation ──────────────────────────────
library(corrplot)
library(pROC)
library(gridExtra)

# ── Class-imbalance handling ────────────────────────────────
library(smotefamily)

set.seed(42)
```

```{r load-data, include=FALSE}

df <- read.csv("fetal_health.csv")

# Convert target variable to a labelled factor 
# 1 = Normal | 2 = Suspect | 3 = Pathological
df$fetal_health <- factor(df$fetal_health,
                           levels = c(1, 2, 3),
                           labels = c("Normal", "Suspect", "Pathological"))


# Shared colour palette
colors_class <- c(
  "Normal" = "#2ecc71",
  "Suspect" = "#f39c12",
  "Pathological" = "#e74c3c"
)
```

# Data and Experiment Setup

## Dataset Overview

```{r data-overview}
cat("Dataset dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
cat("Missing values:", sum(is.na(df)), "\n")
```

```{r class-distribution}
class_dist <- df %>%
  count(fetal_health, name = "n") %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))

kable(
  class_dist,
  col.names = c("Health Class", "Sample Count", "Percentage (%)"),
  caption = "Table 1. Fetal Health Class Distribution"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

```{r class-distribution-plot, fig.height=5}
ggplot(class_dist, aes(x = fetal_health, y = n, fill = fetal_health)) +
  geom_col(width = 0.6, color = "white", linewidth = 1) +
  geom_text(
    aes(label = paste0(n, "\n(", Percentage, "%)")),
    vjust = -0.3,
    fontface = "bold",
    size = 4
  ) +
  scale_fill_manual(values = colors_class) +
  labs(
    title = "Figure 1. Fetal Health Class Distribution",
    x = NULL,
    y = "Sample Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  ) +
  ylim(0, max(class_dist$n) * 1.15)
```

The dataset is strongly imbalanced: the Normal class makes up most observations, while Pathological cases represent a much smaller share of the data. Because minority classes are clinically important, model selection in this report is based primarily on macro-F1 rather than accuracy alone.

## Descriptive Statistics

The features below were selected based on clinical relevance and the strongest separation seen during exploratory analysis.

```{r descriptive-stats}
key_vars <- c(
  "baseline.value",
  "accelerations",
  "prolongued_decelerations",
  "abnormal_short_term_variability",
  "mean_value_of_short_term_variability",
  "histogram_mean"
)

desc_stats <- df %>%
  select(all_of(key_vars)) %>%
  summarise(
    across(
      everything(),
      list(
        Min = ~ round(min(.), 2),
        Mean = ~ round(mean(.), 2),
        Max = ~ round(max(.), 2),
        SD = ~ round(sd(.), 2)
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = c("Feature", ".value"),
    names_sep = "_(?=[^_]+$)"
  )

kable(
  desc_stats,
  caption = "Table 2. Descriptive Statistics for Selected Predictors"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

## Feature Correlation Heatmap

```{r correlation, fig.width=10, fig.height=9}
features <- df %>% select(-fetal_health)
cor_matrix <- cor(features)

corrplot(
  cor_matrix,
  method = "color",
  type = "lower",
  tl.cex = 0.7,
  tl.col = "black",
  addCoef.col = "black",
  number.cex = 0.5,
  col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  title = "Figure 2. Feature Correlation Matrix",
  mar = c(0, 0, 2, 0)
)
```

## Key Feature Distributions by Health Class

```{r feature-density, fig.height=9, fig.width=12}
key_features <- c(
  "baseline.value",
  "accelerations",
  "prolongued_decelerations",
  "abnormal_short_term_variability",
  "mean_value_of_short_term_variability",
  "percentage_of_time_with_abnormal_long_term_variability"
)

density_plots <- lapply(key_features, function(feat) {
  ggplot(df, aes(x = .data[[feat]], fill = fetal_health)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = colors_class) +
    labs(
      title = gsub("[_.]", " ", feat) %>% str_to_title(),
      x = NULL,
      y = "Density"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = if (feat == key_features[1]) "right" else "none",
      plot.title = element_text(face = "bold", size = 9)
    )
})

grid.arrange(
  grobs = density_plots,
  ncol = 3,
  top = grid::textGrob(
    "Figure 3. Feature Density Distributions by Health Class",
    gp = grid::gpar(fontface = "bold", fontsize = 14)
  )
)
```

```{r feature-boxplots, fig.height=9, fig.width=12}
box_plots <- lapply(key_features, function(feat) {
  ggplot(df, aes(x = fetal_health, y = .data[[feat]], fill = fetal_health)) +
    geom_boxplot(alpha = 0.7, outlier.size = 1) +
    scale_fill_manual(values = colors_class) +
    labs(
      title = gsub("[_.]", " ", feat) %>% str_to_title(),
      x = NULL,
      y = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", size = 9)
    )
})

grid.arrange(
  grobs = box_plots,
  ncol = 3,
  top = grid::textGrob(
    "Figure 4. Feature Box Plots by Health Class",
    gp = grid::gpar(fontface = "bold", fontsize = 14)
  )
)
```

Exploratory analysis suggests that `prolongued_decelerations` and `abnormal_short_term_variability` provide the clearest separation between Pathological and non-Pathological cases. In contrast, `accelerations` and `mean_value_of_short_term_variability` tend to be lower in the Pathological group.

## Preprocessing Pipeline

```{r train-test-split}
train_idx <- createDataPartition(df$fetal_health, p = 0.8, list = FALSE)
train_data <- df[train_idx, ]
test_data <- df[-train_idx, ]

cat("Training set:", nrow(train_data), "| Test set:", nrow(test_data), "\n")
```

```{r preprocessing}
X_train <- train_data %>% select(-fetal_health)
y_train <- train_data$fetal_health
X_test <- test_data %>% select(-fetal_health)
y_test <- test_data$fetal_health
classes <- levels(df$fetal_health)

# Standardise using training data only to avoid leakage
preproc <- preProcess(X_train, method = c("center", "scale"))
X_train_sc <- predict(preproc, X_train)
X_test_sc <- predict(preproc, X_test)

# Apply SMOTE to the training data only
train_smote_df <- cbind(X_train_sc, fetal_health = as.numeric(y_train))
smote_result <- SMOTE(
  train_smote_df[, -ncol(train_smote_df)],
  target = train_smote_df$fetal_health,
  K = 5,
  dup_size = 0
)

X_train_sm <- smote_result$data %>% select(-class)
y_train_sm <- factor(
  smote_result$data$class,
  levels = c(1, 2, 3),
  labels = c("Normal", "Suspect", "Pathological")
)

train_sm_df <- cbind(X_train_sm, fetal_health = y_train_sm)

cat("Training set size after SMOTE:", nrow(train_sm_df), "\n")
print(table(train_sm_df$fetal_health))
```

The workflow above intentionally applies scaling and SMOTE only to the training data. This prevents data leakage and preserves the hold-out test set as an unbiased estimate of model generalization.

```{r cv-control}
ctrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)
```

# Results

```{r helpers}
compute_macro_f1 <- function(pred, actual, class_labels) {
  f1s <- sapply(class_labels, function(cls) {
    tp <- sum(pred == cls & actual == cls)
    fp <- sum(pred == cls & actual != cls)
    fn <- sum(pred != cls & actual == cls)

    precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
    recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))

    ifelse(
      precision + recall == 0,
      0,
      2 * precision * recall / (precision + recall)
    )
  })

  mean(f1s)
}

compute_macro_auc <- function(prob_df, actual, class_labels) {
  aucs <- sapply(class_labels, function(cls) {
    y_bin <- ifelse(actual == cls, 1, 0)

    if (length(unique(y_bin)) < 2) {
      return(NA_real_)
    }

    as.numeric(auc(roc(y_bin, prob_df[[cls]], quiet = TRUE)))
  })

  mean(aucs, na.rm = TRUE)
}

compute_class_metrics <- function(pred, actual, class_labels) {
  bind_rows(lapply(class_labels, function(cls) {
    tp <- sum(pred == cls & actual == cls)
    fp <- sum(pred == cls & actual != cls)
    fn <- sum(pred != cls & actual == cls)

    precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
    recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))
    f1 <- ifelse(precision + recall == 0, 0, 2 * precision * recall / (precision + recall))

    tibble(
      Class = cls,
      Support = sum(actual == cls),
      Precision = precision,
      Recall = recall,
      F1 = f1
    )
  }))
}

evaluate_model <- function(model_obj, model_name, newdata, y_true, class_labels) {
  pred <- predict(model_obj, newdata = newdata)
  prob <- predict(model_obj, newdata = newdata, type = "prob")
  cv_scores <- model_obj$resample$Mean_F1

  summary_tbl <- tibble(
    Model = model_name,
    Accuracy = mean(pred == y_true),
    F1_Macro = compute_macro_f1(pred, y_true, class_labels),
    AUC_OVR = compute_macro_auc(prob, y_true, class_labels),
    CV_F1 = mean(cv_scores, na.rm = TRUE),
    CV_F1_SD = sd(cv_scores, na.rm = TRUE)
  )

  list(summary = summary_tbl, pred = pred, prob = prob)
}

extract_best_tune <- function(model_obj, model_name) {
  bt <- model_obj$bestTune

  if (is.null(bt) || ncol(bt) == 0) {
    return(tibble(Model = model_name, Best_Hyperparameters = "Default caret settings"))
  }

  tibble(
    Model = model_name,
    Best_Hyperparameters = paste(
      paste(names(bt), unlist(bt[1, ]), sep = " = "),
      collapse = ", "
    )
  )
}
```

## Model Training

```{r fit-knn}
knn_grid <- expand.grid(k = seq(3, 21, by = 2))

knn_model <- train(
  fetal_health ~ ., data = train_sm_df,
  method = "knn",
  trControl = ctrl,
  tuneGrid = knn_grid,
  metric = "Mean_F1"
)
```

```{r fit-multinom}
lr_model <- train(
  fetal_health ~ ., data = train_sm_df,
  method = "multinom",
  trControl = ctrl,
  trace = FALSE,
  MaxNWts = 5000,
  metric = "Mean_F1"
)
```

```{r fit-rf}
rf_grid <- expand.grid(mtry = c(3, 5, 7, 9, 11))

rf_model <- train(
  fetal_health ~ ., data = train_sm_df,
  method = "rf",
  trControl = ctrl,
  tuneGrid = rf_grid,
  ntree = 300,
  metric = "Mean_F1"
)
```

```{r fit-gbm}
gbm_grid <- expand.grid(
  n.trees = c(100, 300, 500),
  interaction.depth = c(1, 3, 5),
  shrinkage = c(0.01, 0.05, 0.1),
  n.minobsinnode = c(10)
)

gbm_model <- train(
  fetal_health ~ ., data = train_sm_df,
  method = "gbm",
  trControl = ctrl,
  tuneGrid = gbm_grid,
  verbose = FALSE,
  metric = "Mean_F1"
)
```

```{r fit-nnet}
nn_grid <- expand.grid(
  size = c(5, 10, 15),
  decay = c(0, 0.001, 0.01, 0.1)
)

nn_model <- train(
  fetal_health ~ ., data = train_sm_df,
  method = "nnet",
  trControl = ctrl,
  tuneGrid = nn_grid,
  maxit = 500,
  MaxNWts = 5000,
  trace = FALSE,
  metric = "Mean_F1"
)
```

## Best Hyperparameters

```{r best-hyperparameters}
best_tuning <- bind_rows(
  extract_best_tune(knn_model, "KNN"),
  extract_best_tune(lr_model, "Multinomial Logistic Regression"),
  extract_best_tune(rf_model, "Random Forest"),
  extract_best_tune(gbm_model, "GBM"),
  extract_best_tune(nn_model, "Neural Network")
)

kable(
  best_tuning,
  caption = "Table 3. Best Hyperparameters Selected by Repeated Cross-Validation",
  col.names = c("Model", "Selected Setting")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

## Calculate evaluation metrics using saved predictions
```{r}
make_result_from_existing <- function(model_name, pred, prob, y_true, classes, model_obj) {
  accuracy <- mean(pred == y_true)

  f1_per_class <- sapply(classes, function(cls) {
    tp <- sum(pred == cls & y_true == cls)
    fp <- sum(pred == cls & y_true != cls)
    fn <- sum(pred != cls & y_true == cls)

    precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
    recall <- ifelse(tp + fn == 0, 0, tp / (tp + fn))

    ifelse(
      precision + recall == 0,
      0,
      2 * precision * recall / (precision + recall)
    )
  })

  f1_macro <- mean(f1_per_class)
  auc_value <- as.numeric(multiclass.roc(y_true, prob)$auc)

  fold_f1 <- model_obj$resample$Mean_F1
  cv_f1 <- mean(fold_f1, na.rm = TRUE)
  cv_f1_sd <- sd(fold_f1, na.rm = TRUE)

  data.frame(
    Model = model_name,
    Accuracy = round(accuracy, 4),
    F1_Macro = round(f1_macro, 4),
    AUC_OVR = round(auc_value, 4),
    CV_F1 = round(cv_f1, 4),
    CV_F1_SD = round(cv_f1_sd, 4)
  )
}
```


# Generate predictions and predicted probabilities for all models.
```{r}
knn_pred <- predict(knn_model, newdata = X_test_sc)
knn_prob <- predict(knn_model, newdata = X_test_sc, type = "prob")

lr_pred <- predict(lr_model, newdata = X_test_sc)
lr_prob <- predict(lr_model, newdata = X_test_sc, type = "prob")

rf_pred <- predict(rf_model, newdata = X_test_sc)
rf_prob <- predict(rf_model, newdata = X_test_sc, type = "prob")

gbm_pred <- predict(gbm_model, newdata = X_test_sc)
gbm_prob <- predict(gbm_model, newdata = X_test_sc, type = "prob")

nn_pred <- predict(nn_model, newdata = X_test_sc)
nn_prob <- predict(nn_model, newdata = X_test_sc, type = "prob")
```




# Summarize each model performance
```{r}
knn_results <- make_result_from_existing(
  "KNN",
  knn_pred,
  knn_prob,
  y_test,
  classes,
  knn_model
)

lr_results <- make_result_from_existing(
  "Multinomial Logistic Regression",
  lr_pred,
  lr_prob,
  y_test,
  classes,
  lr_model
)

rf_results <- make_result_from_existing(
  "Random Forest",
  rf_pred,
  rf_prob,
  y_test,
  classes,
  rf_model
)

gbm_results <- make_result_from_existing(
  "GBM",
  gbm_pred,
  gbm_prob,
  y_test,
  classes,
  gbm_model
)

nn_results <- make_result_from_existing(
  "Neural Network",
  nn_pred,
  nn_prob,
  y_test,
  classes,
  nn_model
)
```

# Combine all model results and sort them by Macro F1
```{r}
all_results <- bind_rows(
  knn_results,
  lr_results,
  rf_results,
  gbm_results,
  nn_results
) %>%
  arrange(desc(F1_Macro))

all_results
```

# Formatting the final model comparison table 
```{r}
kable(
  all_results,
  caption = "Table 4. Cross-Validation and Hold-Out Test Performance Across Models",
  col.names = c("Model", "Accuracy", "Macro F1", "Macro AUC", "CV Mean F1", "CV F1 SD")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(1, bold = TRUE, background = "#d5f5e3")
```

The strongest hold-out macro-F1 was achieved by **`r all_results$Model[1]`** (`r all_results$F1_Macro[1]`), while accuracy remained high across several models because the Normal class dominates the dataset. This is exactly why macro-F1 is more informative than accuracy for the current task.

```{r model-comparison-plot}
sd_lookup <- all_results %>% select(Model, CV_F1_SD)

all_long <- all_results %>%
  pivot_longer(
    cols = c(Accuracy, F1_Macro, AUC_OVR, CV_F1),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    SD = ifelse(
      Metric == "CV_F1",
      sd_lookup$CV_F1_SD[match(Model, sd_lookup$Model)],
      0
    )
  )

ggplot(all_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(
    aes(ymin = Value - SD, ymax = Value + SD),
    position = position_dodge(width = 0.7),
    width = 0.2,
    linewidth = 0.6,
    na.rm = TRUE
  ) +
  geom_text(
    aes(label = round(Value, 3)),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 2.5
  ) +
  ylim(0, 1.12) +
  labs(
    title = "Figure 5. Model Performance Comparison",
    x = NULL,
    y = "Score",
    fill = "Model"
  ) +
  theme_minimal(base_size = 12)
```

## Best Model Breakdown by Class

```{r best-model-class-metrics}
best_model_name <- all_results$Model[1]
best_pred <- model_store[[best_model_name]]$pred
best_prob <- model_store[[best_model_name]]$prob

best_class_metrics <- compute_class_metrics(best_pred, y_test, classes) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(
  best_class_metrics,
  caption = paste0("Table 5. Per-Class Performance for the Best Model (", best_model_name, ")")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

This table is important because strong overall accuracy can still hide weak performance on minority classes. Reporting precision, recall, and F1 by class makes it easier to evaluate whether the best model is actually learning Suspect and Pathological patterns rather than simply predicting Normal most of the time.

## Confusion Matrices for the Top Three Models

```{r confusion-matrices, fig.width=12, fig.height=4}
plot_cm <- function(pred, actual, title) {
  cm <- confusionMatrix(pred, actual)
  cm_df <- as.data.frame(cm$table)

  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = Freq), size = 4.5, fontface = "bold") +
    scale_fill_gradient(low = "white", high = "#3498db") +
    labs(
      title = paste0(title, "\nAccuracy = ", round(cm$overall["Accuracy"], 3)),
      x = "Actual",
      y = "Predicted"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", size = 10),
      axis.text.x = element_text(angle = 30, hjust = 1)
    )
}

top_models <- all_results %>%
  slice_max(F1_Macro, n = 3, with_ties = FALSE) %>%
  pull(Model)

cm_plots <- lapply(top_models, function(model_name) {
  plot_cm(model_store[[model_name]]$pred, y_test, model_name)
})

do.call(
  grid.arrange,
  c(
    cm_plots,
    list(
      ncol = 3,
      top = grid::textGrob(
        "Figure 6. Confusion Matrices for the Top Three Models",
        gp = grid::gpar(fontface = "bold", fontsize = 13)
      )
    )
  )
)
```

## One-vs-Rest ROC Curves

```{r roc-curves, fig.width=12, fig.height=5}
plot_roc_class <- function(class_name) {
  y_bin <- ifelse(y_test == class_name, 1, 0)

  roc_knn <- roc(y_bin, model_store[["KNN"]]$prob[[class_name]], quiet = TRUE)
  roc_lr <- roc(y_bin, model_store[["Multinomial Logistic Regression"]]$prob[[class_name]], quiet = TRUE)
  roc_rf <- roc(y_bin, model_store[["Random Forest"]]$prob[[class_name]], quiet = TRUE)
  roc_gbm <- roc(y_bin, model_store[["GBM"]]$prob[[class_name]], quiet = TRUE)
  roc_nn <- roc(y_bin, model_store[["Neural Network"]]$prob[[class_name]], quiet = TRUE)

  plot(roc_knn, col = "#3498db", lwd = 2,
       main = paste("ROC —", class_name), legacy.axes = TRUE)
  lines(roc_lr, col = "#2ecc71", lwd = 2)
  lines(roc_rf, col = "#e74c3c", lwd = 2)
  lines(roc_gbm, col = "#9b59b6", lwd = 2)
  lines(roc_nn, col = "#e67e22", lwd = 2)
  abline(a = 0, b = 1, lty = 2, col = "gray")

  legend(
    "bottomright",
    cex = 0.65,
    legend = c(
      sprintf("KNN (%.3f)", auc(roc_knn)),
      sprintf("MLR (%.3f)", auc(roc_lr)),
      sprintf("RF  (%.3f)", auc(roc_rf)),
      sprintf("GBM (%.3f)", auc(roc_gbm)),
      sprintf("NN  (%.3f)", auc(roc_nn))
    ),
    col = c("#3498db", "#2ecc71", "#e74c3c", "#9b59b6", "#e67e22"),
    lwd = 2
  )
}

par(mfrow = c(1, 3))
plot_roc_class("Normal")
plot_roc_class("Suspect")
plot_roc_class("Pathological")
par(mfrow = c(1, 1))
```

## Feature Importance from Random Forest

```{r feature-importance, fig.width=8, fig.height=5}
imp <- varImp(rf_model)$importance
imp_df <- data.frame(Feature = rownames(imp), Importance = imp$Overall) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 10)

ggplot(
  imp_df,
  aes(x = reorder(Feature, Importance), y = Importance, fill = Importance)
) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#aed6f1", high = "#1a5276") +
  coord_flip() +
  labs(
    title = "Figure 7. Top 10 Feature Importances from Random Forest",
    x = "Feature",
    y = "Importance Score"
  ) +
  theme_minimal(base_size = 12)
```

These importance scores are useful for predictive interpretation, but they should not be interpreted as causal effects. They show which variables helped the Random Forest separate classes most effectively under the current modeling pipeline.
